EuPPollNet: A European database of plant-pollinator networks

### 2024-06-14 Manuscript/EuPPollNet_Ms.qmd
updating authors

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. Large-scale analyses across multiple studies can quantify patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023] that cannot be examined in a single study. Although macro-ecological approaches make significant contributions to knowledge, such synthesis work must consider variation across studies in the spatio-temporal nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. This emphasizes the importance of providing data in its rawest possible form in datasets that will be utilized for synthesis and macro-ecological studies.

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists [@reverte2023], along with occurrence data [@zattara2021], is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding biodiversity change [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate plant-pollinator interactions at a European level, which will guide research efforts, conservation planning and will set a foundation for future global change research. For example, only over a dozen of European plant-pollinator networks are included in *Mangal*, while *GloBI* focuses on pairwise interactions disconnected from the community context. Assembling and curating the existing information on EU plant-pollinator networks will guide research efforts, conservation planning and will set a foundation for future global change research.

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on plant-pollinator interactions at the European level. The pollinator taxonomic groups include the main orders of entomofauna that visit and pollinate flowering plants in Europe. These comprise insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], accounting for almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examined the taxonomic and sampling coverage of the different plant and pollinator species at the European level with the help of the most up-to-date species checklists and rarefaction analyses. The overall number of potential pollinators and flowering plants in Europe was estimated with the help of the different checklists and sampling coverage. Finally, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

We investigated how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across latitudes and bioclimatic regions. Networks that are more nested than expected by chance exhibit greater redundancy in their ecological functions and more stability in the face of species loss. EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of nestedness, which has only been evaluated with a relatively small number of networks [@bascompte2003; @payrato2019]. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

Europe contains approximately 5,000 species of pollinators and 25,000 species of plants that rely on animal pollination. EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinators groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). The presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also does not to follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

@@

Bees are responsible for the majority of the interactions at the metaweb level, but their relevance for network topology changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only common floral visitors in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. The proportion of honey bees in networks across Europe is higher than in natural communities across the world [~13%\; @hung2018], potentially reflecting the dominance of pasture habitats in EuPPollNet, the highly generalised nature of honeybees, the important role that honeybees are playing as pollinators in Europe, and/or because beekeeping is widely practice in Europe [@steffan2000; @magrach2017; @herrera2020].

>>>>

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. Large-scale analyses across multiple studies can quantify patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023] that cannot be examined in a single study. Although macro-ecological approaches that use ecological interactions make significant contributions to knowledge [@windsor2023], such synthesis work must consider variation across studies in the spatio-temporal nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. This emphasizes the importance of providing data in its rawest possible form in datasets that will be utilized for synthesis and macro-ecological studies.

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists [@reverte2023], along with occurrence data [@zattara2021], is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding biodiversity change [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate plant-pollinator interactions at a European level. For example, only over a dozen of European plant-pollinator networks are included in *Mangal*, while *GloBI* focuses on pairwise interactions disconnected from the community context. Assembling and curating the existing information on EU plant-pollinator networks will guide research efforts, conservation planning and will set a foundation for future global change research.

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on plant-pollinator interactions at the European level. The pollinator taxonomic groups include the main orders of entomofauna that visit and pollinate flowering plants in Europe. These comprise insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], accounting for almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examined the taxonomic and sampling coverage of the different plant and pollinator species at the European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

EuPPollNet contains one of the largest sets of plant-pollinator networks collated to date, providing a unique opportunity to examine the prevalence of key structural metrics across networks. For example, despite the large theoretical literature on the meaning of a nested structure in plant-pollinator networks [@bascompte2007; @guimaraes2020], where generalist species interact with all other species but specialists interact preferentially only with generalists, this pattern has only been empirically evaluated with a relatively small number of networks [@bascompte2003; @staniczenko2013; @payrato2019], and is still debated how structural metrics such as connectance and nestedness change across latitudes and bioclimatic regions [@olesen2002; @trojelsgaard2013; @song2017].

Overall, EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

Europe hosts approximately 5,000 species of pollinators and 25,000 species of plants that rely on animal pollination. EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinators groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). The presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also does not to follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

#Filter out 2 bioregions that only have a small study with data of a single day
data = data %>% 
filter(!Bioregion == "Steppic") %>% 
filter(!Bioregion == "Pannonian")

@@

Bees are responsible for the majority of the interactions at the metaweb level, but their relevance for network topology changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only common floral visitors in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. The proportion of honey bees in networks across Europe is higher than in natural communities across the world [~13%\; @hung2018], potentially reflecting the dominance of intensive grasslands habitats in EuPPollNet, the highly generalised nature of honeybees, the important role that honeybees are playing as pollinators in Europe, and/or because beekeeping is widely practiced in Europe [@steffan2000; @magrach2017; @herrera2020].
